One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2019-12-10 paper-v2.tex
add discussion

In this section we introduce the statistical model for the animal locations and observation process, including a detailed description of how this model can be represented as a modified Poisson likelihood.  We use this formulation to fit the model using the \texttt{R-INLA} implementation of INLA but this likelihood could in principle be used in any statistical software capable of specifying such a modified Poisson form.

@@

spatio-temporal extension. AR1 through time etc.  Citations for this.

covariates on detection function.  High dimensional taylor falls down at some point?

This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.  Three such realizations are in \autoref{fig:intensity-realizations}.  There are some notable differences when compared to the mean and quantile summary maps.  Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.  The summary plots hide the spatial structure the model has been able to estimate from the data.  In general, the mean of a random field will contain less spatial structure than realisations of the field.

We note that this presentation of abundance estimates differs from typical presentations.  It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.  However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.  We find the distinction between expected and realised abundance to be a useful one. Here the posterior intensity process is used to generate possible realisations of the intensity, which are then used to generate the abundance estimates.

>>>>

In this section we introduce the statistical model for the animal locations and observation process, including a detailed description of how this model can be represented as a modified Poisson likelihood.  We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA.

@@

There are several natural extensions to the model.  For this analysis we only considered a single survey year.  Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.  \texttt{R-INLA} has the ability to fit space-time effects that can be represented as a kronecker product of the relevant precision matrices.  An example of this type of space-time interaction effect would be the spatial SPDE above interacting with an AR1 temporal process.  \cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.

Another extension is to consider more complicated observation processes.  For example, we did not use any explanatory covariates for the detection function.  This remains to be investigated further.  Any additional parameters to the detection function would result in a higher-dimensional outer optimisation problem for iterated INLA.  This could lead to problem if, for example, identifiability is an issue.

We note also that our presentation of abundance estimates differs from typical presentations.  It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.  However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.  We find the distinction between expected and realised abundance to be a useful one. The credible intervals for the posterior expected abundance will be narrower than the credible intervals for the posterior realised abundance.  The Bayesian perspective is useful here as well as for the uncertainty in the detection model estimates.  Since outputs based on sampling from the joint posterior of all model parameters, including the detection model, the Bayesian framework means the resulting model outputs we present here naturally average over the uncertainty in the observation process.  [[Not really sure if this stuff is worth mentioning or if can be said in a better way]]

Our use of \texttt{inlabru} and \texttt{R-INLA} for inference means we can take advantage of the sparsity of the GMRF precision matrix, a key advantage of the SPDE approach.  A dense covariance matrix of dimension $n \times n$ typically has an $\mathcal{O}(n^3)$ cost to invert.  In contrast to this \texttt{R-INLA} makes use of sparse matrix algorithms to invert a sparse 2-dimensional precision matrix at a typical cost of $\mathcal{O}(n^{3/2})$ \citep{lindgren_explicit_2011}.  This computational efficienty, in addition to the flexibility of the mesh triangulation, can make the SPDE an attractive choice for implementing high-resolution spatial effects in complex domains.  An additional benefit, as described in detail in \cite{yuan_point_2017}, is that the SPDE can also be implemented on the sphere, making analysis of global datasets computationally feasible.

In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.  Our example threshold was chosen arbitrarily and it is not obvious to us that there will always be an easy or simple choice of thresholds and levels of uncertainty for a given analysis.  This requires the input of relevant domain expertise and statistical knowledge so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.  The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.  Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions.  In the above we chose to use excursion sets, excursion functions and the posterior of realised abundance as a demonstration of this.  It is not always the case that a map of predicted mean abundance will be the best output for a species distribution model.

The point process perspective with spatial SPDE effect can seem fundamentally different to a two-stage GAM approach.  However, we think that it can be useful to highlight the similarities in approach.  It can be tempting to think that because with the point process model does not require us to bin the data into counts that we can avoid some loss of information.  However, we also had to make the same assumption of constant intensity within each transect that is common in more traditional distance sampling approaches.  This means the exactly location of the points is not used to inform the posterior intensity.  The assumption of constant intensity within each point transect is essentially the point process version of binning point data into counts.

The SPDE effect can also seem fundamentally different to, say, a spline-based smoother.  However, there is a fundamental connection between SPDEs and optimal smoothing splines.  The optimal smoothing spline is a lminimum variance unbiased linear estimator of the posterior mean of an SPDE \citep{kimeldorf_spline_2002}.  This means a smoothing spline can be interpreted as an SPDE and vice-versa.  Depending on the methods a user is most familiar with this can be a useful perspective.  \cite{yue_bayesian_2014} show how to implement common spline penalties in an SPDE framework and \cite{miller_understanding_2019} offer a complementary explanation to interpret and fit SPDE models within standard penalised regression splines software.
